Efficiently stable presentations from error-correcting codes

### 2023-10-05 efficient_stability.tex
cleared a comment
some small edits

\author[1]{Michael Chapman}
\author[2]{Thomas Vidick}
\author[3]{Henry Yuen}
\affil[1]{}
\affil[2]{}
\affil[3]{}

@@

A finer property which concerns us here is the \emph{soundness} of the code, a parameter that is connected to the notion of \emph{local testability}. A code can be (non-uniquely) specified through a \emph{parity-check matrix} $h\in\F^{m\times n}$ as $\code = \ker h$. The rows of $h$ are thought of as constraints (``parity checks'') that specify $\code$ as a subspace of $\F^n$. A code is called locally testable \emph{with soundness $\rho$} if for every $x\in \F^n$, $\frac{1}{m}|hx|\geq \rho \, \frac{1}{n}\, d(x,\code)$ \hnote{added the divide by $n$}, where $d(x,\code)$ denotes the minimum of $|x-c|$ over $c\in \code$.  Ideally one would like to design families of good codes such that in addition $\rho$ is bounded below by a constant independent of $n$.\footnote{Note that in principle a code can have a small minimal distance $d$, and still be locally testable with soundness $\rho>0$. So a family of codes can be ``locally testable'' without being ``good.''} This is a challenging task, and the construction of families of good locally testable codes was only achieved very recently~\cite{LTC_DELLM}.

@@

\begin{definition}[Stability]\label{def:eff-stab-intro}
Let $G = \langle S:R\rangle $ be a finitely presented group, $\mu_S$ a distribution on $S$ and $\mu_R$ a distribution on $R$. For $\delta:[0,1]\to[0,1]$ such that $\lim_{t\to 0}\delta(t)=0$ and an integer $d\geq 1$ we say that the presentation $G=\langle S:R\rangle$ is $(\delta,\mu_S,\mu_R,d)$-stable if for every $(\eps,\mu_R)$-almost homomorphism $\phi: \mF(S) \to \mU(\C^d)$ there is a unitary representation $\psi: G \to \mU_d(\C)$ \hnote{there's some mismatch, is it $\mU(\C^d)$ or $\mU_d(\C)$?} such that
\[ \Es{s\sim \mu_S} \big\|  \phi(s) - \psi(s) \big \|_f^2 \,\leq\, \delta(\eps)\;.\]
We refer to any function $\delta$ satisfying the above as a ``modulus of stability'' of the presentation.
\end{definition}

@@

In this paper we also consider a version of \emph{flexible} stability, i.e.\ the representation $\psi$ is allowed to range in $U_{d'}(\C)$ for some $d'\neq d$; see Definition~\ref{def:eff-stab} for the general definition. This requires a more careful definition of closeness; for now we restrict our attention to this simpler definition. % and for the purposes of this introduction we restrict ourselves to the simpler definition.

We can now ask the question: is $G(h)$, the group presentation defined in~\eqref{eq:gh-intro}, stable according to Definition~\ref{def:eff-stab-intro}? One can verify that with $\mu_R$ and $\mu_S$ the uniform distribution on $R$ and $S$ respectively, $G(h)$ is $(\delta,\mu_S,\mu_R,1)$-stable for $\delta=\eps/\rho$ if and only if $\code = \ker h$ is locally testable with soundness $\rho>0$ \hnote{spelled out relationship between $\delta,\rho$}. But what about higher-dimensional approximate representations? Are these also stable, or does one need to make further requirements on $\code$ beyond local testability? We do not yet have a comprehensive answer to these questions. \hnote{elaborated a little on the challenges} However, the answer cannot be straightforward: even deducing basic properties about the group $G(h)$ given its presentation -- let alone its stability properties -- appears to be a challenging task. For example, there are examples of parity check matrices $h$ for which the group $G(h)$ is non-Abelian or even infinite; see Remark~\ref{rk:non-abelian}.

\paragraph{Efficient stability.}
Faced with the apparent difficulty of studying the general question, it is time to refine our focus and formulate the question which we \emph{do} address. To start, let us explicitly note that 
while stability has previously (for the most part) been studied as a question about a group, our formulation makes it a question about a \emph{presentation} of the group. In particular it is known~\cite{gowers2017inverse,de2019operator} that for finite groups $G$ (which are the only groups we consider in this paper) the multiplication table presentation, which has $|G|$ generators for every group element and $|G|^2$ relations for every pairwise product, is $\delta(\eps)=C\eps$-stable\footnote{\hnote{added this:}Here we assume that the $\mu_R,\mu_S$ are uniform over the relations and generators of the multiplication table presentation, respectively. Furthermore, \cite{gowers2017inverse,de2019operator} use a notion of flexible stability, where the nearby exact representation may act on a larger dimension.} for some constant $C$ that is independent of the group. This holds even for approximate representations in arbitrary tracial von Neumann algebras, as long as one allows the notion of flexible stability alluded to above.

@@

\begin{theorem*}[Main, informal]
For every integer $k\geq 1$ there is a presentation $\Z_2^k = \langle S_k:R_k\rangle$ such that $|S_k|,|R_k| = 2^{\poly\log(k)}$ and furthermore this presentation is $(\delta,\mu_{S_k},\mu_{R_k},d)$-stable for all $d \in \N$, where $\delta(\eps)=\poly(\log k,\eps)$, $\mu_{S_k}$ is uniform over the generators and $\mu_{R_k}$ is some distribution over the relations. \hnote{Spelled the conditions out because if we just let $\mu_S,\mu_R$ be anything, then we get a trivial statement if $\mu_S$, say, is just supported on one generator and $\mu_R$ is just on one relation.}
\end{theorem*}

@@

\paragraph{Acknowledgments.}

@@

\begin{lemma}\label{lem:lower_bound_on_stability_rate_standard_presentation_Z_2^k}
    Let $\mu_R$ and $\mu_S$ be as in Lemma~\ref{lem:eff-z2}. % be the uniform distribution over the commutation relations in the presentation $\langle S\colon R\rangle$ from \eqref{eq:z2-efficient}, and $\mu_S$ the marginal distribution from $\mu_R$, which is the uniform one on $S=\{x_1,...,x_k\}$. 
    There is a $(\nicefrac{1}{\binom{k}{2}},\mu_R)$-almost homomorphism of $\langle S\colon R\rangle$ which is at least $\nicefrac{1}{16k}$ far from any homomorphism. Namely, the modulus of stability of \eqref{eq:z2-efficient} is $\Omega(k\eps)$.\tnote{Is the `namely'' part immediate? Meaning, why could it not be $\Omega(\sqrt{\eps})$, which for large $\eps$ is better? Maybe you can get an example for a larger range of $\eps$ --- here $\eps$ depends on the size of the presentation. But maybe we can also just leave it at the first sentence.}\mnote{You are right. I meant that there is no better linear bound... Is there a succinct way of saying that? :)}
\end{lemma}

@@

\begin{remark}
As was discussed in the introduction \mnote{Here we should refer to the discussion on the difference between $L^1$ and $L^\infty$ definitions},  it is more common to use the $L^\infty$ analogue of stability. A slight modification of the example given in  the proof of Lemma~\ref{lem:lower_bound_on_stability_rate_standard_presentation_Z_2^k} provides a lower bound also in that setup. The idea is as follows: In the construction we provided, the pair $(1,2)$ played a special role, since $A_1$ was a CNOT conditioned on the second bit. We can thus define the analogous construction for every pair $(i,j)$ where $1\leq i<j\leq k$. By tensoring all of these constructions along the pairs $(i,j)$, we get $k$ permutations acting on $\binom{[k]}{2}\times \field^{k+1}$, which in turn provide a lower bound in the $L^\infty$ setup. The proof is similar to the above, and since this is a slight digression from our main theme, we omit it from the text. 
\end{remark}

@@

\begin{lemma}\label{lem:eff-pauli}
  Let $\mC$ be the class of tracial von Neumann algebra. Let $\mu_R$ be the equal mixture of the uniform distribution on all words $ [x_i,x_j]$, $[z_i,z_j]$, $[x_i,z_j]$ ($i\neq j$), the uniform distibution over all $[x_i,z_i]$, $[x_i,J]$, $[z_i,J]$, $x_i^2$ and $z_i^2$, and $J^2$. Let $\mu_S$ be the marginal of $\mu_R$, as described in Remark~\ref{rk-mus}. Then 
 the presentation~\eqref{eq:z2-efficient} is $(O(k\eps),\mu_S,\mu_R,\mC)$-stable.
\hnote{what's the dependence on $k$?}
\end{lemma}

@@

\begin{remark}\label{rk:non-abelian}
There exists matrices $h$ such that $G(h)$ is not $\Z_2^k$, and in fact is not Abelian. For an example, see~\cite[Example 2.16]{paddock2022arkhipov}. For that example, $n=12$, $k=7$, and the parity check matrix $h$ can be described explicitly as follows: the $7$ rows of $h$ are indexed by the vertices of the complete bipartite graph $K_{3,4}$, the $12$ columns are indexed by the $12$ edges of $K_{3,4}$, and the entry $(i,j)$ of $h$ is $1$ if and only if the edge $j$ is incident on vertex $i$. As shown in~\cite{paddock2022arkhipov}, $G(h)$ is not Abelian. By considering $K_{3,6}$ instead of $K_{3,4}$, one in addition obtains a non-Abelian \emph{infinite} group. 
\end{remark}

>>>>

\author[1]{Michael Chapman}
\author[2]{Thomas Vidick}
\author[3]{Henry Yuen}
\affil[1]{}
\affil[2]{ Faculty of Mathematics and Computer Science, Weizmann Institute of Science, Israel}
\affil[3]{}

@@

A finer property which concerns us here is the \emph{soundness} of the code, a parameter that is connected to the notion of \emph{local testability}. A code can be (non-uniquely) specified through a \emph{parity-check matrix} $h\in\F^{m\times n}$ as $\code = \ker h$. The rows of $h$ are thought of as constraints (``parity checks'') that specify $\code$ as a subspace of $\F^n$. A code is called locally testable \emph{with soundness $\rho$} if for every $x\in \F^n$, $\frac{1}{m}|hx|\geq \rho \, \frac{1}{n}\, d(x,\code)$, where $d(x,\code)$ denotes the minimum of $|x-c|$ over $c\in \code$.  Ideally one would like to design families of good codes such that in addition $\rho$ is bounded below by a constant independent of $n$.\footnote{Note that in principle a code can have a small minimal distance $d$, and still be locally testable with soundness $\rho>0$. So a family of codes can be ``locally testable'' without being ``good.''} This is a challenging task, and the construction of families of good locally testable codes was only achieved very recently~\cite{LTC_DELLM}.

@@

\begin{definition}[Stability]\label{def:eff-stab-intro}
Let $G = \langle S:R\rangle $ be a finitely presented group, $\mu_S$ a distribution on $S$ and $\mu_R$ a distribution on $R$. For $\delta:[0,1]\to[0,1]$ such that $\lim_{t\to 0}\delta(t)=0$ and an integer $d\geq 1$ we say that the presentation $G=\langle S:R\rangle$ is $(\delta,\mu_S,\mu_R,d)$-stable if for every $(\eps,\mu_R)$-almost homomorphism $\phi: \mF(S) \to \mU(\C^d)$ there is a unitary representation $\psi: G \to \mU(\C^d)$ \hnote{there's some mismatch, is it $\mU(\C^d)$ or $\mU_d(\C)$?}\tnote{Let's use the former, I made the correction here} such that
\[ \Es{s\sim \mu_S} \big\|  \phi(s) - \psi(s) \big \|_f^2 \,\leq\, \delta(\eps)\;.\]
We refer to any function $\delta$ satisfying the above as a ``modulus of stability'' of the presentation.
\end{definition}

@@

In this paper we also consider a version of \emph{flexible} stability, i.e.\ the representation $\psi$ is allowed to range in $U_{d'}(\C)$ for some $d'\neq d$; see Definition~\ref{def:eff-stab} for the general definition. This requires a more careful definition of closeness; for now we restrict our attention to the simpler definition. % and for the purposes of this introduction we restrict ourselves to the simpler definition.

We can now ask the question: is $G(h)$, the group presentation defined in~\eqref{eq:gh-intro}, stable according to Definition~\ref{def:eff-stab-intro}? One can verify that with $\mu_R$ and $\mu_S$ the uniform distribution on $R$ and $S$ respectively, $G(h)$ is $(\delta,\mu_S,\mu_R,1)$-stable for $\delta=\eps/\rho$ if and only if $\code = \ker h$ is locally testable with soundness $\rho>0$. But what about higher-dimensional approximate representations? Are these also stable, or does one need to make further requirements on $\code$ beyond local testability? We do not yet have a comprehensive answer to these questions. However, the answer cannot be straightforward: even deducing basic properties about the group $G(h)$ given its presentation -- let alone its stability properties -- appears to be a challenging task. For example, there are examples of parity check matrices $h$ for which the group $G(h)$ is non-Abelian or even non-amenable; see Remark~\ref{rk:non-abelian}.

\paragraph{Efficient stability.}
Faced with the apparent difficulty of studying the general question, it is time to refine our focus and formulate the question which we \emph{do} address. To start, let us explicitly note that 
while stability has previously (for the most part) been studied as a question about a group, our formulation makes it a question about a \emph{presentation} of the group. In particular it is known~\cite{gowers2017inverse,de2019operator} that for finite groups $G$ (which are the only groups we consider in this paper) the multiplication table presentation, which has $|G|$ generators for every group element and $|G|^2$ relations for every pairwise product, is $\delta(\eps)=C\eps$-stable\footnote{Here we assume that $\mu_R,\mu_S$ are uniform over the relations and generators of the multiplication table presentation, respectively. Furthermore, the results of~\cite{gowers2017inverse,de2019operator} apply to a notion of \emph{strongly flexible} stability, where the nearby exact representation may act on a space of larger, but not too much larger, dimension.} for some constant $C$ that is independent of the group. This holds even for approximate representations in arbitrary tracial von Neumann algebras, as long as one allows the notion of flexible stability alluded to above.

@@

\begin{theorem*}[Main, informal]
For every integer $k\geq 1$ there is a presentation $\Z_2^k = \langle S_k:R_k\rangle$ such that $|S_k|,|R_k| = 2^{\poly\log(k)}$ and furthermore this presentation is $(\delta,\mu_{S_k},\mu_{R_k},d)$-stable for all $d \in \N$, where $\delta(\eps)=\poly(\log k,\eps)$, $\mu_{S_k}$ is uniform over the generators and $\mu_{R_k}$ is some distribution over the relations. %\hnote{Spelled the conditions out because if we just let $\mu_S,\mu_R$ be anything, then we get a trivial statement if $\mu_S$, say, is just supported on one generator and $\mu_R$ is just on one relation.}
\end{theorem*}

@@

\paragraph{Acknowledgments.} TV is supported by a research grant from the Center for New Scientists at the Weizmann Institute of Science, a Simons Investigator award, AFOSR Grant No. FA9550-22-1-0391, and ERC Consolidator Grant VerNisQDevS (101086733).

@@

\begin{lemma}\label{lem:lower_bound_on_stability_rate_standard_presentation_Z_2^k}
    Let $\mu_R$ and $\mu_S$ be as in Lemma~\ref{lem:eff-z2}. % be the uniform distribution over the commutation relations in the presentation $\langle S\colon R\rangle$ from \eqref{eq:z2-efficient}, and $\mu_S$ the marginal distribution from $\mu_R$, which is the uniform one on $S=\{x_1,...,x_k\}$. 
    There is a $(\nicefrac{1}{\binom{k}{2}},\mu_R)$-almost homomorphism of $\langle S\colon R\rangle$ which not $(\nicefrac{1}{16k},\mu_S)$-close to any homomorphism (according to Definition~\ref{def:close}). 
\end{lemma}

\tnote{added this here to clarify:}
Lemma~\ref{lem:lower_bound_on_stability_rate_standard_presentation_Z_2^k} implies that the modulus of stability of \eqref{eq:z2-efficient} is $\Omega(k\eps)$ as $\eps\to 0$. This is because for any $\eps>0$ such that $\eps\leq 1/\binom{k}{2}$ we can create an $(\eps,\mu_R)$-almost homomorphism of $\langle S\colon R\rangle$ by taking direct sums of one copy of the almost-homomorphism $\phi$ from Lemma~\ref{lem:lower_bound_on_stability_rate_standard_presentation_Z_2^k}, which ranges into $\mU(\C^{2^{k+1}})$, with $R=\lceil 2^{k+1}/(\binom{k}{2}\eps)\rceil$ copies of the trivial representation of $G(h)$. Clearly, a $\delta$-close homomorphism implies, by restriction, a $\delta R$-close homomorphism to $\phi$. Hence $\delta R \geq 1/(16k)$, which gives $\delta=\Omega(k\eps)$ as claimed.

@@

\begin{remark}
As discussed in the introduction, in the literature on stability it is more common to use a $L^\infty$ analogue of our Definition~\ref{def:eff-stab}, where the notion of almost-homomorphism and closeness are both measured by taking a supremum over relations and generators respectively, as opposed to averaging according to distributions $\mu_R,\mu_S$. 
A slight modification of the example given in  the proof of Lemma~\ref{lem:lower_bound_on_stability_rate_standard_presentation_Z_2^k} provides a lower bound also in that setup. The idea is as follows: In the construction we provided, the pair $(1,2)$ played a special role, since $A_1$ was a CNOT conditioned on the second bit. We can thus define the analogous construction for every pair $(i,j)$ where $1\leq i<j\leq k$. By tensoring all of these constructions along the pairs $(i,j)$, we get $k$ permutations acting on $\binom{[k]}{2}\times \field^{k+1}$, which in turn provide a lower bound in the $L^\infty$ setup. The proof is similar to the above, and since this is a slight digression from our main theme, we omit the details. 
\end{remark}

@@

\begin{lemma}\label{lem:eff-pauli}
  Let $\mC$ be the class of tracial von Neumann algebra. Let $\mu_R$ be the equal mixture of the uniform distribution on all words $ [x_i,x_j]$, $[z_i,z_j]$, $[x_i,z_j]$ ($i\neq j$), the uniform distibution over all $[x_i,z_i]$, $[x_i,J]$, $[z_i,J]$, $x_i^2$ and $z_i^2$, and $J^2$. Let $\mu_S$ be the marginal of $\mu_R$, as described in Remark~\ref{rk-mus}. Then 
 the presentation~\eqref{eq:z2-efficient} is $(O(k\eps),\mu_S,\mu_R,\mC)$-stable.
\end{lemma}

@@

\begin{remark}\label{rk:non-abelian}
There exists matrices $h$ such that $G(h)$ is not $\Z_2^k$, and in fact is not Abelian. For an example, see~\cite[Example 2.16]{paddock2022arkhipov}. For that example, $n=12$, $k=7$, and the parity check matrix $h$ can be described explicitly as follows: the $7$ rows of $h$ are indexed by the vertices of the complete bipartite graph $K_{3,4}$, the $12$ columns are indexed by the $12$ edges of $K_{3,4}$, and the entry $(i,j)$ of $h$ is $1$ if and only if the edge $j$ is incident on vertex $i$. As shown in~\cite{paddock2022arkhipov}, $G(h)$ is not Abelian. By considering $K_{3,6}$ instead of $K_{3,4}$, one in addition obtains a non-Abelian \emph{infinite} group. Using similar arguments it is possible to construct $h$ such that $G(h)$ is not amenable, etc.; see the discussion in~\cite[Section 6]{paddock2022arkhipov}.
\end{remark}
